Unlocking Sensitivity for Visibility-based Estimators of the 21\,cm Reionization Power Spectrum

### 2018-02-15 Rotation paper/draft_working.tex
removing boldtexts

Visibility-based pipelines are not fundamentally limited to using only fully redundant baselines. In fact, most sensitivity forecasts to date do include partial redundancy \citep{Pobersens, HERA, JoshAntPos}. \textbf{A number of methods to produce power spectrum measurements using partially redundant baselines has been studied in recent years. A power-spectrum estimator based on visibilities gridded in the $uv$-plane was introduced in \cite{Choudhury2014Estimators, Choudhury2016Taperingb, Choudhury2016Taperinga}, and compared to a simple pair-baseline estimator in \cite{Choudhury2014Estimators}. Due to considerations of computational complexity, the authors favored grid-based estimator over pair-baseline estimators despite the higher accuracy of the latter.  \cite{wterm} proposed another approach based on gridded visibilities in sky-tracking measurements. 
%Arrays capable of tracking include MWA and LOFAR, but not drift-scan telescopes such as PAPER, HERA, and potentially SKA-low.
\cite{CTrott} studied three observing strategies (tracking, drift scan, and a combination of the two) and developed a pair-visibility based approach to power spectrum estimation.
%and concluded that zenith drift scan was the preferred observing strategy.
In this paper, we extend the above works and introduce a non-gridded \emph{baseline}-pair power spectrum estimator that can be applied to both tracking-capable arrays and drift scan-only arrays. Our formalism also provides a way to identify which baseline pairs to cross-correlate, overcoming the aforementioned computational disadvantage of baseline-based estimators relative to grid-based estimators.
Furthermore, compared to
%the pair-baseline estimators studied in
\cite{Choudhury2014Estimators} and \cite{CTrott}, our analysis 
more explicitly treats the curved nature of the sky as well as its spectral dimension.
%of the sensitivities indicate the importance of the three dimensional nature of the sky, resulting in us directly addressing curved sky effects in our formalism. 
}

@@

We shall use the 128-element PAPER array (henceforth referred to as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Section \ref{sec:arrconf}. 
The PAPER array \textbf{was} located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern is shown
in Fig. \ref{fig:AntPos}. The array \textbf{consisted} of a 112-element core in a rectangular grid, and 16 ``outriggers'' used primarily to aid calibration.  In the bottom panel, the two baselines marked in red are an example of an equivalent pair. We denote a equivalency class of baselines in the PAPER grid by their separations, in this case  \{2,0\}, for the
antennas are separated by 2 units east and 0 units north. Similarly,
the baselines marked in blue and yellow are respectively examples
of \{2,1\} and $\{2, -1\}$.
Note that \{2,0\} and $\{-2, 0\}$ are the same class of baselines and should not be included twice in power-spectrum measurements. Antennas in purely north-south baselines
are close (4m), and hence these baselines are not suitable
for sensitive measurements due to cross-coupling. On the other hand, the small North-South separation means that classes such as \{2,0\} and  \{2,1\} are expected to be near-equivalent. The PAPER-64 analysis of \cite{Ali2015} \textbf{used} three classes of baselines, the PAPER-128
equivalent of which \textbf{were} 
\{2, 0\}, \{2, 1\} and $\{2, -1\}$. There, baselines \textbf{were} cross-multiplied within each class. This paper provides the method for inter-class multiplications. We will use the short hand notation $\{m,n:m',n'\}$ to denote a pair of baseline classes to be cross-multiplied.

@@

With implicit bounds of integrals from $-\infty$ to $\infty$, we have
\begin{eqnarray}
 &&\langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau)\rangle \nonumber \\
 && =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\langle T^{*}(\boldsymbol{r})T(\boldsymbol{r'})\rangle B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}) \nonumber \\
 &&  =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}\frac{d^{3}\kappa'}{(2\pi)^{3}}\langle T^{*}(\boldsymbol{\kappa})T(\boldsymbol{\kappa'})\rangle e^{-i(\boldsymbol{\kappa}\cdot \boldsymbol{r}-\boldsymbol{\kappa'}\cdot\boldsymbol{r'})}\right) \nonumber \\
 && \qquad  \times B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'})\nonumber \\
 && =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}P(\kappa)e^{-i\boldsymbol{\kappa}\cdot(\boldsymbol{r}-\boldsymbol{r'})}\right) \nonumber \\
 && \qquad \times B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}),
 \end{eqnarray}
 where 
 \begin{eqnarray}
\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'})\equiv &\exp\left[i\frac{2\pi}{c}\left(\boldsymbol{b}\cdot\nu_{r}\hat{\boldsymbol{r}}-\boldsymbol{b'}\cdot\nu_{r'}\boldsymbol{\Gamma}\hat{\boldsymbol{r}}'\right)\right] \nonumber \\
& \times \exp\left[-i2\pi\tau\left(\nu_{r}-\nu_{r'}\right)-i\psi_{\nu}\right].
\end{eqnarray}
The third equality of Eq.(\ref{eq:main}) follows from assuming the translational invariance of the statistics of the 21\,cm field. Continuing, we may make the assumption that the 3D power spectrum varies negligibly over the $k$-space of interest, thus allowing us to pull out of the integral the power spectrum centered at 
\[
k_{\boldsymbol{b}, \tau} \equiv 2\pi \sqrt{\left(\frac{\tau}{Y}\right)^2 + \left(\frac{\bar{b}}{\lambda X}\right)^2},
\label{eq:kbtau}
\]
where $\bar{b}$ is approximately the mean of $b$ and $b'$. This gives 
 \begin{eqnarray}
  &&\langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau)\rangle \nonumber \\
 && \approx P(k_{\bar{b}, \tau})\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2} \delta_{D}^{(3)}(\boldsymbol{r}-\boldsymbol{r'})B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'}) \Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}) \nonumber \\
 && = P(k_{\bar{b}, \tau})\!\!\int\!\! \frac{d^{3}r}{(X^2Y)^2}B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r}) e^{-i2\pi\frac{\nu_{r}}{c}\left(\hat{\boldsymbol{r}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma} \hat{\boldsymbol{r}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}}, \qquad
\label{eq:main}
\end{eqnarray}
where $\delta_D$ is the Dirac delta-function, and notice that the phase factor $\exp\left[-i2\pi\tau\left(\nu-\nu'\right)\right]$
drops out in the end. In factoring out the power spectrum from the integral, one is essentially making the approximation that the estimator probes only a single Fourier mode. In Section \ref{sec:chromaticity} we will relax this assumption and examine the exact form of the Fourier-space footprint that is being probed, as well as its effect on foreground isolation.

@@

One sees that all four window functions are fairly localized around specific values of $k$, which to a good approximation are described by Eq. \eqref{eq:kbtau}. This suggests that both equivalent- and near-equivalent-baseline delay-spectrum estimators are good estimators of the power spectrum. \textbf{For window functions that are centered on higher $\ell$ modes, the window functions for both estimators become elongated in $k$, with stronger tails towards the low $k$ modes. Since the low $k$ modes are where the foregrounds reside, measurements of the high $\ell$ modes therefore mix in more foreground power.} This is the phenomenology of the wedge, where at fine angular scales the foreground leakage to higher $k$ modes are more pronounced. Importantly, however, we note that the window functions for the near-equivalent-baseline estimator are no more elongated than for the equivalent-baseline estimator. One may thus conclude that our proposed near-equivalent-baseline estimator does not result in extra foreground contamination of the EoR window.

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time offset $\Delta t_{\boldsymbol{bb'}}$ (in hours),  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$ (in meters). \textbf{Shown on the diagonal are histograms of given quantities for the different arrays. On the off-diagonals we show the the scatter plots of the three quantities against each other.} Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta t_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. \textbf{We show the histograms on the diagonals and scatter plots on the off-diagonals.} We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. \textbf{As seen in the scatter plot of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$ in Fig. \ref{fig:pairplot}, these are also the shorter baselines, which tend to contribute sensitivity to lower $k$-modes.} The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. \textbf{However, compared to PAPER, HERA's boost in sensitivity is mostly attributable to the longer baselines---and thus high $k$---modes (see Fig. \ref{fig:pairplot}). } The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}

>>>>

Visibility-based pipelines are not fundamentally limited to using only fully redundant baselines. In fact, most sensitivity forecasts to date do include partial redundancy \citep{Pobersens, HERA, JoshAntPos}. A number of methods to produce power spectrum measurements using partially redundant baselines has been studied in recent years. A power-spectrum estimator based on visibilities gridded in the $uv$-plane was introduced in \cite{Choudhury2014Estimators, Choudhury2016Taperingb, Choudhury2016Taperinga}, and compared to a simple pair-baseline estimator in \cite{Choudhury2014Estimators}. Due to considerations of computational complexity, the authors favored grid-based estimator over pair-baseline estimators despite the higher accuracy of the latter.  \cite{wterm} proposed another approach based on gridded visibilities in sky-tracking measurements. 
%Arrays capable of tracking include MWA and LOFAR, but not drift-scan telescopes such as PAPER, HERA, and potentially SKA-low.
\cite{CTrott} studied three observing strategies (tracking, drift scan, and a combination of the two) and developed a pair-visibility based approach to coherently combine visibilities gridded on the $uv$-plane for power spectrum estimation.
%and concluded that zenith drift scan was the preferred observing strategy.
In this paper, we extend the above works and introduce a non-gridded \emph{baseline}-pair power spectrum estimator that can be applied to both tracking-capable arrays and drift scan-only arrays. Our formalism also provides a way to identify which baseline pairs to cross-correlate, overcoming the aforementioned computational disadvantage of baseline-based estimators relative to grid-based estimators.
Furthermore, compared to
%the pair-baseline estimators studied in
\cite{Choudhury2014Estimators} and \cite{CTrott}, our analysis 
more explicitly treats the curved nature of the sky as well as its spectral dimension.
%of the sensitivities indicate the importance of the three dimensional nature of the sky, resulting in us directly addressing curved sky effects in our formalism.

@@

We shall use the 128-element PAPER array (henceforth referred to as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Section \ref{sec:arrconf}. 
The PAPER array was located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern is shown
in Fig. \ref{fig:AntPos}. The array consisted of a 112-element core in a rectangular grid, and 16 ``outriggers'' used primarily to aid calibration.  In the bottom panel, the two baselines marked in red are an example of an equivalent pair. We denote a equivalency class of baselines in the PAPER grid by their separations, in this case  \{2,0\}, for the
antennas are separated by 2 units east and 0 units north. Similarly,
the baselines marked in blue and yellow are respectively examples
of \{2,1\} and $\{2, -1\}$.
Note that \{2,0\} and $\{-2, 0\}$ are the same class of baselines and should not be included twice in power-spectrum measurements. Antennas in purely north-south baselines
are close (4m), and hence these baselines are not suitable
for sensitive measurements due to cross-coupling. On the other hand, the small North-South separation means that classes such as \{2,0\} and  \{2,1\} are expected to be near-equivalent. The PAPER-64 analysis of \cite{Ali2015} used three classes of baselines, the PAPER-128
equivalent of which were
\{2, 0\}, \{2, 1\} and $\{2, -1\}$. There, baselines were cross-multiplied within each class. This paper provides the method for inter-class multiplications. We will use the short hand notation $\{m,n:m',n'\}$ to denote a pair of baseline classes to be cross-multiplied.

@@

With implicit bounds of integrals from $-\infty$ to $\infty$, we have
\begin{eqnarray}
 &&\langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau)\rangle \nonumber \\
 && =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\langle T^{*}(\boldsymbol{r})T(\boldsymbol{r'})\rangle B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}) \nonumber \\
 &&  =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}\frac{d^{3}\kappa'}{(2\pi)^{3}}\langle T^{*}(\boldsymbol{\kappa})T(\boldsymbol{\kappa'})\rangle e^{-i(\boldsymbol{\kappa}\cdot \boldsymbol{r}-\boldsymbol{\kappa'}\cdot\boldsymbol{r'})}\right) \nonumber \\
 && \qquad  \times B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'})\nonumber \\
 && =\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2}\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}P(\kappa)e^{-i\boldsymbol{\kappa}\cdot(\boldsymbol{r}-\boldsymbol{r'})}\right) \nonumber \\
 && \qquad \times B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'})\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}),
 \label{eq:main0}
 \end{eqnarray}
 where 
 \begin{eqnarray}
\Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'})\equiv &\exp\left[i\frac{2\pi}{c}\left(\boldsymbol{b}\cdot\nu_{r}\hat{\boldsymbol{r}}-\boldsymbol{b'}\cdot\nu_{r'}\boldsymbol{\Gamma}\hat{\boldsymbol{r}}'\right)\right] \nonumber \\
& \times \exp\left[-i2\pi\tau\left(\nu_{r}-\nu_{r'}\right)-i\psi_{\nu}\right].
\end{eqnarray}
The third equality of Eq. (\ref{eq:main0}) follows from assuming the translational invariance of the statistics of the 21\,cm field. Continuing, we may make the assumption that the 3D power spectrum varies negligibly over the $k$-space of interest, thus allowing us to pull out of the integral the power spectrum centered at 
\[
k_{\boldsymbol{b}, \tau} \equiv 2\pi \sqrt{\left(\frac{\tau}{Y}\right)^2 + \left(\frac{\bar{b}}{\lambda X}\right)^2},
\label{eq:kbtau}
\]
where $\bar{b}$ is approximately the mean of $b$ and $b'$. This gives 
 \begin{eqnarray}
  &&\langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau)\rangle \nonumber \\
 && \approx P(k_{\bar{b}, \tau})\int \frac{d^{3}rd^{3}r'}{(X^2Y)^2} \delta_{D}^{(3)}(\boldsymbol{r}-\boldsymbol{r'})B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r'}) \Phi(\boldsymbol{r},\boldsymbol{\Gamma} \boldsymbol{r'}) \nonumber \\
 && = P(k_{\bar{b}, \tau})\!\!\int\!\! \frac{d^{3}r}{(X^2Y)^2}B^{*}(\boldsymbol{r})B(\boldsymbol{\Gamma} \boldsymbol{r}) e^{-i2\pi\frac{\nu_{r}}{c}\left(\hat{\boldsymbol{r}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma} \hat{\boldsymbol{r}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}}, \qquad
\label{eq:main}
\end{eqnarray}
where $\delta_D$ is the Dirac delta-function, and notice that the phase factor $\exp\left[-i2\pi\tau\left(\nu-\nu'\right)\right]$
drops out in the end. In factoring out the power spectrum from the integral, one is essentially making the approximation that the estimator probes only a single Fourier mode. In Section \ref{sec:chromaticity} we will relax this assumption and examine the exact form of the Fourier-space footprint that is being probed, as well as its effect on foreground isolation.

@@

One sees that all four window functions are fairly localized around specific values of $k$, which to a good approximation are described by Eq. \eqref{eq:kbtau}. This suggests that both equivalent- and near-equivalent-baseline delay-spectrum estimators are good estimators of the power spectrum. For window functions that are centered on higher $\ell$ modes, the window functions for both estimators become elongated in $k$, with stronger tails towards the low $k$ modes. Since the low $k$ modes are where the foregrounds reside, measurements of the high $\ell$ modes therefore mix in more foreground power. This is the phenomenology of the wedge, where at fine angular scales the foreground leakage to higher $k$ modes are more pronounced. Importantly, however, we note that the window functions for the near-equivalent-baseline estimator are no more elongated than for the equivalent-baseline estimator. One may thus conclude that our proposed near-equivalent-baseline estimator does not result in extra foreground contamination of the EoR window.

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time offset $\Delta t_{\boldsymbol{bb'}}$ (in hours),  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$ (in meters). On the diagonals we show histograms of given quantities for the different arrays, with the vertical axes linear in scale. On the off-diagonals we show the the scatter plots of the three quantities against each other. Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta t_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. We show the histograms on the diagonals and scatter plots on the off-diagonals. We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. As seen in the scatter plot of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$ in Fig. \ref{fig:pairplot}, these are also the shorter baselines, which tend to contribute sensitivity to lower $k$-modes. The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. However, compared to PAPER, HERA's boost in sensitivity is mostly attributable to the longer baselines---and thus high $k$-modes (see Fig. \ref{fig:pairplot}). The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}
